---
title: "Polisci 270 Final Cheat Sheet"
author: "Fayzaan Virk"
date: "2026-05-01"
output: pdf_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
library(ps270data)
```

```{r include=F, eval=F}
data(package = "ps270data")
```


## Data Wrangling

Data wrangling is the process of taking a complicated data set and making it easier to work with. This can mean filtering or selecting only a certain number or rows or columns, adding new variables, or pivoting the table. In R, we use the group_by(), summarize(), pivot_wider(), mutate(), filter(), select(), pivot_longer(), and the three join() commands to wrangle our data.

I will use the gapminder data set to demonstrate some of these commands

### group_by() and summarize()

First, I can use the group_by() and summarize() commands to find the mean GDP per capita for each continent.

```{r}
gapminder |>
  group_by(continent) |>
  summarize(mean_gdppc = mean(gdpPercap))
```

### filter() and select()

I can use the filter() command to only see certain rows of the data set.

```{r}
gapminder |>
  filter(year == 2007)
```

The select() command works the same way as filter() but for choosing certain columns

```{r}
gapminder |>
  filter(year == 2007) |>
  select(country, gdpPercap)
```

### mutate() and rename()

The mutate() command can create new columns using logicals or calculations and rename() can change existing columns' names. 

```{r}
gapminder |>
  mutate(
    gdp = gdpPercap * pop,
    post_coldwar = if_else(year > 1991, 1, 0)
  ) |>
  rename("population" = pop)
```

### arrange()

The arrange() function will reorder the entries in a column

```{r}
gapminder |>
  filter(year == 2007) |>
  arrange(desc(gdpPercap))
```


Add in a section for join() and pivot_longer()

See Code from Lecture 4, 5, 9, 12, Problem Set 2, 3, 4, 5, 6, 8

## Average Treatment Effect

The average treatment effect is a way of calculation how much a treatment affects the outcome variable. The formula is the expected outcome variable if all subjects were treated minus the expected outcome if all subjects were not treated. To approximate this, we use the mean of the treatment group for the outcome variable minus the mean of the control group.

I'll use the trains experiments (Spanish-speakers in train stations in Boston).

```{r}
trains |>
  group_by(treatment) |>
  summarize(post_mean = mean(numberim.post)) |>
  pivot_wider(
    names_from = treatment,
    values_from = post_mean
  ) |>
  mutate(
    ate = `1` - `0`
  )
```

Additionally, you can use select() and pull() to get just the numeric value

```{r}
trains |>
  group_by(treatment) |>
  summarize(post_mean = mean(numberim.post)) |>
  pivot_wider(
    names_from = treatment,
    values_from = post_mean
  ) |>
  mutate(
    ate = `1` - `0`
  ) |>
  select(ate) |>
  pull()
```


See Code from Lecture 7, Problem Set 3, 7

## Ggplot

ggplot2 is a package that is part of tidyverse that is used to create many types of graphs and visual representations of data. In class, we learned scatterplots, box and whisker plots, barplots, and histograms. These are created using geom_point(), geom_box(), geom_bar() or geom_col(), and geom_histogram(). For a scatterplot, a line of best fit can be added using geom_smooth().

### Scatterplot

```{r}
midwest |>
  ggplot(aes(x = percblack, y = percbelowpoverty)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = F) +
  labs(
    x = "Percentage of black population",
    y = "Percentage below the poverty line",
    title = "Relationship between black and impoverished population in Midwestern counties"
  )
```

### Bar plots

```{r}
midwest |>
  group_by(state) |>
  summarize(population = sum(poptotal)) |>
  ggplot(aes(x = state, y = population)) +
  geom_col() + 
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "State",
    y = "Total Population",
    title = "Population in each Midwestern State"
  )
```

### Histogram

```{r}
midwest |>
  ggplot(aes(x = percbelowpoverty)) +
  geom_histogram(binwidth = 2) + 
  labs(
    x = "Percentage Below the Poverty Line",
    y = "Count",
    title = "Percentage Below the Poverty Line in Midwestern Counties"
  )
```

### Box and Whisker Plot

```{r}
midwest |>
  ggplot(aes(x = percbelowpoverty)) +
  geom_boxplot() +
  facet_wrap(~ state) +
  scale_y_discrete() +
  labs(
    x = "Percentage Below the Poverty Line",
    title = "Percentage Below the Poverty Line in Midwestern Counties by State"
  )
```

### Colors

```{r}
midwest |>
  ggplot(
    aes(
      x = percbelowpoverty,
      fill = state,
      color = state
    )) +
  geom_density(alpha = 0.3) +
  labs(
    x = "Percentage Below the Poverty Line",
    y = "Density",
    title = "Percentage Below the Poverty Line in Midwestern Counties by State"
  )
```


See Code from Lecture 3, Problem Set 1, 2, 3, 4, 5, 6, 7

## Experimental Types

An observational study is where we simply observe how two variables interact in real life. We do not assign a treatment. This is very common in real life, since we often can't assign treatments due to ethical or logistical reasons. Because we cannot randomly assign treatments, it is much harder to establish causality from an observational study because of a greater chance of confounding variables. On the other hand, Randomized Controlled Trials (RCTs) do have randomly assigned treatment groups. RCTs tend to have a higher internal validity, meaning it is easier to conclude a causal relationship from them. Observational studies tend to have a higher external validity, which means the conclusion can be more easily generalized beyond this study.

### Research Designs

There are three types of research designs. A cross-sectional design compares the control and treatment groups after the treatment has been administered. This assumes the treatment and control groups are on average identical. A before-and-after design compares the treatment group from before and after the treatment was administered. This assumes none of the confounders will vary with time. A difference-in-differences design used the before-and-after of the control group to infer what would've happened to the treatment group if no treatment was administered. This assumes that the trends that affect one group equally affect the other group. All three rely on assumptions that can't always be verified, and are thus more susceptible to confounding variables than a RCT. There are some ways to try and control for confounding variables using statistical procedures, one of which is subclassification. Subclassification is where treated and control groups are compared within the levels of a confounder.

See Code from Lecture 8, Problem Set 3

## Proportion Tables

```{r}
cces_2020 |>
  group_by(pres_vote) |>
  summarize(n = n()) |>
  mutate(proportion = n / sum(n))
```

For multiple grouping variables, see below

```{r}
cces_2020 |>
  filter(pres_vote %in% c("Joe Biden (Democrat)", "Donald J. Trump (Republican)")) |>
  mutate(pres_vote = if_else(pres_vote == "Joe Biden (Democrat)", "Biden", "Trump")) |>
  group_by(pid3, pres_vote) |>
  summarize(n = n()) |>
  mutate(prop = n/sum(n)) |>
  select(!n) |>
  pivot_wider(
    names_from = pid3,
    values_from = prop
  ) |>
  knitr::kable()
```


See Code from Lecture 9, 10

## Z-Score

The z-score represent how many standard deviations away from the mean a data point is. The formula is (x - mu)/sigma. 

See Code from Lecture 11

## Linear Models

See Code from Lecture 14, 15, 16, 25, Problem Set 5

## Bootstrapping

See Code from Lecture 17, 19, Problem Set 6, 7, 8

## Confidence Intervals (Code)

See Code from Lecture 20, Problem Set 7

## Confidence Intervals (Math)

See Problem Set 7

## Hypothesis Testing (Code)

See Code from Lecture 21, 22, 25, Problem Set 8

## Hypothesis Testing (Math)

See Problem Set 8

## Normal Distributions

See Code from Lecture 23















