---
title: "Polisci 270 Final Cheat Sheet"
author: "Fayzaan Virk"
date: "2026-05-01"
output: pdf_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
library(ps270data)
```

```{r include=F, eval=F}
data(package = "ps270data")
```


## Data Wrangling

Data wrangling is the process of taking a complicated data set and making it easier to work with. This can mean filtering or selecting only a certain number or rows or columns, adding new variables, or pivoting the table. In R, we use the group_by(), summarize(), pivot_wider(), mutate(), filter(), select(), pivot_longer(), and the three join() commands to wrangle our data.

I will use the gapminder data set to demonstrate some of these commands

### group_by() and summarize()

First, I can use the group_by() and summarize() commands to find the mean GDP per capita for each continent.

```{r}
gapminder |>
  group_by(continent) |>
  summarize(mean_gdppc = mean(gdpPercap))
```

### filter() and select()

I can use the filter() command to only see certain rows of the data set.

```{r}
gapminder |>
  filter(year == 2007)
```

The select() command works the same way as filter() but for choosing certain columns

```{r}
gapminder |>
  filter(year == 2007) |>
  select(country, gdpPercap)
```

### mutate() and rename()

The mutate() command can create new columns using logicals or calculations and rename() can change existing columns' names. 

```{r}
gapminder |>
  mutate(
    gdp = gdpPercap * pop,
    post_coldwar = if_else(year > 1991, 1, 0)
  ) |>
  rename("population" = pop)
```

### arrange()

The arrange() function will reorder the entries in a column

```{r}
gapminder |>
  filter(year == 2007) |>
  arrange(desc(gdpPercap))
```


Add in a section for join() and pivot_longer()

See Code from Lecture 4, 5, 9, 12, Problem Set 2, 3, 4, 5, 6, 8

## Average Treatment Effect

The average treatment effect is a way of calculation how much a treatment affects the outcome variable. The formula is the expected outcome variable if all subjects were treated minus the expected outcome if all subjects were not treated. To approximate this, we use the mean of the treatment group for the outcome variable minus the mean of the control group.

I'll use the trains experiments (spanish-speakers in train stations in Boston).

```{r}
trains |>
  group_by(treatment) |>
  summarize(post_mean = mean(numberim.post)) |>
  pivot_wider(
    names_from = treatment,
    values_from = post_mean
  ) |>
  mutate(
    ate = `1` - `0`
  )
```

Additionally, you can use select() and pull() to get just the numeric value

```{r}
trains |>
  group_by(treatment) |>
  summarize(post_mean = mean(numberim.post)) |>
  pivot_wider(
    names_from = treatment,
    values_from = post_mean
  ) |>
  mutate(
    ate = `1` - `0`
  ) |>
  select(ate) |>
  pull()
```


See Code from Lecture 7, Problem Set 3, 7

## Ggplot

See Code from Lectures 3, Problem Set 1, 2, 3, 4, 5, 6, 7

## Experimental Types

See Code from Lecture 8, Problem Set 3

## Proportion Tables

See Code from Lecture 9, 10

## Z-Score

See Code from Lecture 11

## Linear Models

See Code from Lecture 14, 15, 16, 25, Problem Set 5

## Bootstrapping

See Code from Lecture 17, 19, Problem Set 6, 7, 8

## Confidence Intervals (Code)

See Code from Lecture 20, Problem Set 7

## Confidence Intervals (Math)

See Problem Set 7

## Hypothesis Testing (Code)

See Code from Lecture 21, 22, 25, Problem Set 8

## Hypothesis Testing (Math)

See Problem Set 8

## Normal Distributions

See Code from Lecture 23















